---
title: "BeeBDC Duplicate Status by Data Source Analysis"
format: 
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    geometry: margin=1in
    fontsize: 11pt
    prefer-html: true
date: today
author: "Gretchen LeBuhn"
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)
library(DT)
```

# Overview

This analysis examines the distribution of duplicate status classifications across different data sources in the Alarcon-Cruz *et al.* dataset.

```{r}
#| label: load-data

# Load data
data <- read.csv("~/GitHub/phylo-endemism/data/BeeDataNoiNat_Clean.csv", stringsAsFactors = FALSE)

cat("Total records in dataset:", format(nrow(data), big.mark = ","), "\n")
cat("Records with source information:", format(sum(!is.na(data$source) & data$source != ""), big.mark = ","), "\n")
cat("Records with duplicate status:", format(sum(!is.na(data$duplicateStatus) & data$duplicateStatus != ""), big.mark = ","), "\n")
```

# Overall Duplicate Status Distribution

```{r}
#| label: overall-duplicate-status

# Overall duplicate status summary
overall_duplicate_summary <- data %>%
  filter(!is.na(duplicateStatus) & duplicateStatus != "") %>%
  count(duplicateStatus, sort = TRUE) %>%
  mutate(percentage = round(n / sum(n) * 100, 1))

kable(overall_duplicate_summary,
      caption = "Overall Distribution of Duplicate Status Classifications",
      col.names = c("Duplicate Status", "Records", "Percentage %"),
      format.args = list(big.mark = ","))
```

# Data Sources Summary

```{r}
#| label: data-sources

# Summary of data sources - unique specimens only
source_summary <- data %>%
  filter(!is.na(source) & source != "" &
         (str_detect(tolower(duplicateStatus), "kept") | 
          str_detect(tolower(duplicateStatus), "unique") |
          duplicateStatus == "" |
          is.na(duplicateStatus))) %>%
  count(source, sort = TRUE) %>%
  mutate(percentage = round(n / sum(n) * 100, 1)) %>%
  rename(unique_specimens = n)

cat("Number of data sources:", nrow(source_summary), "\n")
cat("Total  specimens with source information:", format(sum(source_summary$unique_specimens), big.mark = ","), "\n\n")

kable(source_summary,
      caption = "Specimens by Data Source",
      col.names = c("Data Source", "Unique Specimens", "Percentage %"),
      format.args = list(big.mark = ","))
```

# Duplicate Status by Data Source

```{r}
#| label: duplicate-by-source-table

# Create cross-tabulation of source vs duplicate status
duplicate_by_source <- data %>%
  filter(!is.na(source) & source != "" & 
         !is.na(duplicateStatus) & duplicateStatus != "") %>%
  count(source, duplicateStatus) %>%
  pivot_wider(names_from = duplicateStatus, values_from = n, values_fill = 0)

# Add row totals
duplicate_by_source <- duplicate_by_source %>%
  mutate(Total = rowSums(select(., -source))) %>%
  arrange(desc(Total))

# Display as table
kable(duplicate_by_source,
      caption = "Duplicate Status Counts by Data Source",
      format.args = list(big.mark = ","))
```

# Visualization: Unique Specimens by Data Source

```{r}
#| label: visualization
#| fig-width: 12
#| fig-height: 6

# Get top sources for visualization
top_sources_for_plot <- source_summary %>%
  head(15) %>%
  pull(source)

# Create bar chart for unique specimens by source
plot_data <- source_summary %>%
  head(15) %>%
  mutate(source = reorder(source, unique_specimens))

ggplot(plot_data, aes(x = source, y = unique_specimens)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_text(aes(label = format(unique_specimens, big.mark = ",")), 
            hjust = -0.1, size = 3) +
  coord_flip() +
  labs(
    title = "Specimens by Data Source",
    subtitle = "Top 15 sources (duplicates excluded)",
    x = "Data Source",
    y = "Number of Unique Specimens"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold")
  ) +
  scale_y_continuous(labels = scales::comma)
```

# Institutional Analysis - Only specimen data used in analysis

```{r}
#| label: institution-extraction-unique

# Function to extract institution codes from catalog numbers
extract_institution <- function(catalog_num) {
  if(is.na(catalog_num) || catalog_num == "") return(NA)
  
  # Extract alphabetic prefix before numbers
  prefix <- str_extract(catalog_num, "^[A-Za-z]+")
  return(prefix)
}

# Filter to only unique specimens (exclude duplicates)
unique_specimens <- data %>%
  filter(str_detect(tolower(duplicateStatus), "kept") | 
         str_detect(tolower(duplicateStatus), "unique") |
         duplicateStatus == "" |
         is.na(duplicateStatus))

# Apply institution extraction to unique specimens only
unique_specimens$institution_code <- sapply(unique_specimens$catalogNumber, extract_institution)

# Create institution lookup with common codes
institution_lookup <- data.frame(
  code = c("AMNH", "ANSP", "BBSL", "BMNH", "CAS", "EMEC", "FSCA", "INHS", 
           "KUNHM", "LACM", "MCZ", "MSUC", "NMNH", "OSUC", "PCYU", "SDNHM", 
           "SEMC", "TAMU", "UAIC", "UBCZ", "UCB", "UCDC", "UCMS", "UCRC", 
           "UCD", "USNM", "WIRC", "YPM"),
  institution = c("American Museum of Natural History",
                  "Academy of Natural Sciences of Philadelphia", 
                  "Bee Biology and Systematics Laboratory",
                  "Natural History Museum, London",
                  "California Academy of Sciences",
                  "Essig Museum of Entomology, UC Berkeley",
                  "Florida State Collection of Arthropods",
                  "Illinois Natural History Survey",
                  "Kansas University Natural History Museum",
                  "Los Angeles County Museum",
                  "Museum of Comparative Zoology, Harvard",
                  "Michigan State University",
                  "National Museum of Natural History",
                  "Ohio State University Collection",
                  "Pacific Coast Entomological Society",
                  "San Diego Natural History Museum",
                  "Snow Entomological Museum, Kansas",
                  "Texas A&M University",
                  "University of Arizona Insect Collection",
                  "University of British Columbia",
                  "University of California, Berkeley",
                  "UC Davis Center for Population Biology",
                  "UC Museum of Paleontology",
                  "UC Riverside Entomology Collection",
                  "UC Davis Entomology Collection",
                  "US National Museum",
                  "Wisconsin Insect Research Collection",
                  "Yale Peabody Museum"),
  stringsAsFactors = FALSE
)

# Count unique specimens by institution code
institution_summary_unique <- unique_specimens %>%
  filter(!is.na(institution_code) & institution_code != "") %>%
  count(institution_code, name = "Unique_Specimens") %>%
  left_join(institution_lookup, by = c("institution_code" = "code")) %>%
  mutate(institution = ifelse(is.na(institution), 
                             institution_code,  # Use the code itself instead of "Unknown Institution (code)"
                             institution)) %>%
  arrange(desc(Unique_Specimens))

cat("Total institutions identified from unique specimens:", nrow(institution_summary_unique), "\n")
cat("Unique specimens with institution codes:", 
    format(sum(institution_summary_unique$Unique_Specimens), big.mark = ","), "\n")
cat("Total unique specimens in dataset:", 
    format(nrow(unique_specimens), big.mark = ","), "\n")
```

## Institution Summary Table - Specimens used in analysis only

```{r}
#| label: institution-table-unique

# Create comprehensive institution table for unique specimens
institution_table_unique <- institution_summary_unique %>%
  mutate(Percentage = round(Unique_Specimens / sum(Unique_Specimens) * 100, 1)) %>%
  select(institution_code, institution, Unique_Specimens, Percentage) %>%
  rename("Institution Code" = institution_code,
         "Institution Name" = institution,
         "Number of Specimens" = Unique_Specimens,
         "Percentage %" = Percentage)

# For PDF output, use kable instead of datatable
kable(institution_table_unique,
      caption = "Institutional Sources of Bee Specimens used in analysis (Duplicates Excluded)",
      format.args = list(big.mark = ","))

# Show top 20 for summary
cat("\n\nTop 20 Institutions by Number of Bee Specimens used in analysis (Unique):\n")
kable(head(institution_table_unique, 20),
      caption = "Top 20 Institutions by Number of Bee Specimens used in analysis (Unique)",
      format.args = list(big.mark = ","))
```

```{r}
#| label: conclusions
#| echo: false

total_sources <- nrow(source_summary)
sources_with_duplicates <- length(unique(data$source[str_detect(tolower(data$duplicateStatus), "duplicate") & !is.na(data$source)]))
```
